---
title: "ess_gene_mean"
author: "Chen Dong"
date: "12/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## Preprocess
# Read in essential and non-essential gene expression data
ess_gx <- data.table::fread("./expdata_essential_gene.csv", fill = T)
ess.gx_colname <- colnames(ess_gx)[-length(colnames(ess_gx))]
ess_gx <- ess_gx[,-1]
colnames(ess_gx) <- ess.gx_colname

noness_gx <- data.table::fread("./expdata_nonessential_gene_data.csv", fill = T)
noness.gx_colname <- colnames(noness_gx)[-length(colnames(noness_gx))]
noness_gx <- noness_gx[,-1]
colnames(noness_gx) <- noness.gx_colname

# log transforma the expression data
ess_gx[, 3:ncol(ess_gx)] <- log2(ess_gx[, 3:ncol(ess_gx)]+1)
noness_gx[, 3:ncol(noness_gx)] <- log2(noness_gx[, 3:ncol(noness_gx)]+1)

# remove genes that have zero expression
ess_gx <- as.data.frame(ess_gx)
ess_gx_nonzero <- ess_gx[, 3:ncol(ess_gx)][apply(ess_gx[, 3:ncol(ess_gx)], 2, function(z) !any(z == 0))]
ess_gx <- cbind(SMTS=ess_gx$SMTS,ess_gx_nonzero)
noness_gx <- as.data.frame(noness_gx)
noness_gx_nonzero <- noness_gx[, 3:ncol(noness_gx)][apply(noness_gx[, 3:ncol(noness_gx)], 2, function(z) !any(z == 0))]
noness_gx <- cbind(SMTS=noness_gx$SMTS,noness_gx_nonzero)
```


```{r}
# calculate the gene expression mean for each gene in each tissue
x <- apply(ess_gx[, 2:ncol(ess_gx)],2,function(x) aggregate(x ~ SMTS, data = ess_gx, FUN= "mean" ))
tissue <- x$AAMP$SMTS
# process the data
y <- as.data.frame(lapply(x,function(a) a$x)) 
ess_gx_mean <- cbind(tissue,y)
str(ess_gx_mean)
```

# Structure of the ess_gx_mean
- 30x547
- first col is the tissue types as factors, 30 types in total
- rest cols are essential genes with each element in the col as the mean gene exp in a tissue
- 546 genes in total

```{r}
# calculate the gene expression mean for each gene in each tissue
p <- apply(noness_gx[, 2:ncol(noness_gx)],2,function(x) aggregate(x ~ SMTS, data = noness_gx, FUN= "mean" ))
# process the data
q <- as.data.frame(lapply(p,function(a) a$x)) 
noness_gx_mean <- cbind(tissue,q)
str(noness_gx_mean)
```

# Structure of the noness_gx_mean
- 30x138
- first col is the tissue types as factors, 30 types in total
- rest cols are nonessential genes with each element in the col as the mean gene exp in a tissue
- 137 genes in total
- be aware of the difference in the number of total genes in one tissue if doing KS test

# Please proceed the test below in the same RMD

```{r}
gx_mean_tissue <- data.frame(type = c(rep("essential",ncol(ess_gx_mean) - 1),rep("non essential",ncol(noness_gx_mean)-1)))

ks_results <- c()

for(t in tissue){
  ess_distribution <- as.numeric(ess_gx_mean[which(ess_gx_mean$tissue==t),-1])
  noness_distribution <- as.numeric(noness_gx_mean[which(ess_gx_mean$tissue == t),-1])
  gx_mean_tissue[,t] <- c(ess_distribution,noness_distribution)
  
  pval <- ks.test(ess_distribution, noness_distribution)$p.value
  ks_results <- append(ks_results, pval)
}

ks_results <- data.frame(tissue = tissue, ks_results = ks_results)
```
# The statistic results of Kolmogorov-Smirnov Tests is stored in ks_results
- the results is mean expression of essential genes v.s mean expression of non-essential gens in each tissue


# Plot the density of essential and non-essential genes
```{r}
library(ggplot2)
library(gridExtra)

parseTissue <- gsub(" ","",tissue)
colnames(gx_mean_tissue)[-1] <- parseTissue

gglist <- list()
for(t in parseTissue){
  gglist[[t]] <- ggplot(gx_mean_tissue, aes_string(x = t)) + geom_density(aes(col = type))

}

#ggplot(gx_mean_tissue, aes(x = Brain)) + geom_density(aes(col = type))
#do.call("grid.arrange", c(gglist, ncol= 5))

```

# All the plot is stored in gglist, to retrieve the plot of certain tissue, use "gglist$Tissue", for example : 
```{r}
gglist$Brain
```
